---
title: "Introduction to `data.table`"
author: "Haema Nilakanta & Kim Ky"
date: "10/10/2018"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 10px;
}
h1 { /* Header 1 */
  font-size: 26px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 14px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

___

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
```

# Overview

### What is `data.table`?

"`data.table` is an R package that provides an enhanced version of `data.frame`s." It provides a faster and more efficient way to do data manipulation. 

### Install and load

```{r,eval=F}
# install 
install.packages("data.table")
# load 
library(data.table)
```

# The Basics



# The Advanced 

*Compare base R and data.table here - more advanced operations*

# So, why should you use `data.table`?

* concise and consistent syntax
    + `DT[i, j, by]`
* speed
    + faster than base R and `dplyr`
* efficient memory usage
    + sub-assign by reference: `DT[x < 0, y := NA]`
    + aggregate while joining: `DT1[DT2, list(z=sum(z)), by = .EACHI]` (equivalent to `DT1[, .(z=sum(z)), keyby=.(x,y)][DT2]`)
* features
    + `fread`, `fwrite`, automatic indexing, rolling joins, non-equivalent joins, etc.

More detail, [here](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly).


# References

1. Matt Dowle and Arun Srinivasan (2018). `data.table`: Extension of `data.frame`. R
  package version 1.11.4. https://CRAN.R-project.org/package=data.table
2. https://github.com/Rdatatable/data.table/wiki
3. https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly